`methyvim`: an R package for Variable Importance Measures for

### 2017-10-07 paper/paper_BiocF1000.Rmd
adding the Summary and final touches

---
title: '`methyvim`: an R-package for Variable Importance Measures for
Differential Methylation Analysis'
author:
  - name: Nima S. Hejazi
    affiliation: Address of author1
  - name: Rachael Phillips
    affiliation: orchid.org/0000-0002-8474-591X
  - name: Alan E. Hubbard
    affiliation: Address of author2
  - name: Mark J. van der Laan
    affiliation: Address of author2
abstract:
  We present a general algorithm for the nonparametric estimation of
  treatment/exposure effects on DNA methylation at CpG sites, complete with straightforward statistical inference for such estimates. This approach
  leverages variable importance measures, a class of parameters that arise in
  causal inference and are defined in such a manner that they provide targeted
  estimates of the relative importance of individual CpG site with respect to
  a binary exposure/treatment assigned at the level of subjects. Additionally,
  this procedure is computationally efficient; incorporating a preliminary
  screening step to isolate a subset of sites for which there is cursory
  evidence of differential methylation and a multiple testing correction step
  to control the False Discovery Rate as if all sites were tested. This
  technique for analysis of differentially methylated positions provides an
  avenue to move beyond linear models without the loss of interpretable
  statistical inference.

@@

DNA methylation is a fundamental epigenetic process known to play an important
role in the regulation of gene expression. DNA CpG methylation involves the
addition of a methyl group ($\text{CH}_3$) to the fifth carbon of the cytosine
ring structure to form 5-methylcytosine. Numerous biological and medical studies
have implicated DNA CpG methylation as playing a role in disease and
development [@robertson2005dna]. Perhaps unsurprisingly then, biotechnologies
have been developed to study the molecular mechanisms of this epigenetic
process. Modern assays, like the Illumina _Infinium_ Methylation assay,
allow for quantitative interrogation of DNA methylation of CpG sites scattered
across the genome at single-nucleotide resolution; moreover, much effort
has been invested, by the bioinformatics community, in the development of tools  
for properly removing technological effects that may contaminate biological
signatures measured by such assays. Despite these advances in both biological
and bioninformatical techniques, most statistical methods available for
differential analysis of data produced by such assays rely on over-simplified
(often generalized linear) models. This is because typical statistical
estimation approaches are not achievable for such high-dimensional data without restrictive modeling assumptions. When these assumptions are violated, estimates
of the population-level effect of an exposure or treatment may have large bias.
Data-adaptive estimation procedures that utilize machine learning can control
for high-dimensional confounders when estimating treatment/exposure effects.
However, interpretable statistical inference (i.e. confidence intervals,
hypothesis tests) resulting from these data-adaptive estimates is challenging [@libbrecht2015machine]. In this paper, we present an alternative such
statistical analysis approaches in the form of nonparametric estimation
procedures that provide simple and well interpretable statistical inference.
Inspired by concepts of counterfactual effects from the causal inference
literature and machine learning algorithms, we provide a computationally
efficient technique for obtaining targeted estimates of nonparametric _variable importance measures_ (__VIM__) for a pre-screened set of CpGs, controlling for
the False Discovery Rate (FDR) as if all sites were tested. Under counterfactual assumptions, these estimators have asymptotic normal sampling distributions
and thus provide reliable inference. Counterfactuals are "counter"-fact, defined
in terms of an unrealistic yet ideal experiment. In the context of DNA
methylation and for a binary exposure/treatment, we define the counterfactual
outcomes as the Beta-value/M-value a CpG site would have if all subjects were
exposed/treated and the Beta-value/M-value a CpG site would have if all
subjects were not exposed/treated. Of course, these counterfactual outcomes are
impossible to observe but they are very helpful for us to asses the importance
of a CpG site relative to a subjects treatment/exposure and they can be
statistically estimated [@van2006statistical]. Our VIM definition uses the
concepts of counterfactuals and the framework for estimation is built on
targeted maximum likelihood estimation or targeted minimum loss-based estimation
(TMLE) [@van2006targeted]. This methodology targets the individual importance
of each CpG site by taking advantage of the prediction power of machine learning
algorithms. The mathematical properties of this approach have been explored in [@tuglus2008targeted] and we focus on our contribution to it's application,
`methyvim`, an R package that accommodates this technique for differential
methylation analysis.

@@

The core functionality of this package is made available via the eponymous
`methyvim` function, which implements a statistical algorithm designed to
compute targeted estimates of VIMs, defined in such a way that the VIMs
represent parameters of scientific interest in computational biology
experiments; moreover, these VIMs are defined such that they may be estimated in
a manner that is very nearly assumption-free, that is, within a _fully
nonparametric statistical model_. __The statistical algorithm consists in
several major steps:__

1. Pre-screening of genomic sites is used to isolate a subset of sites for
  which there is cursory evidence of differential methylation. For the sake of
  computational feasibility, targeted minimum loss-based estimates of VIMs are
  computed only for this subset of sites. Currently, the available screening
  approach adapts core routines from the
  [`limma`](http://bioconductor.org/packages/release/bioc/html/limma.html) R
  package. Future releases will support functionality from other
  packages (e.g., [`randomForest`](https://cran.r-project.org/web/packages/randomForest/index.html),
  [`tmle.npvi`](https://CRAN.R-project.org/package=tmle.npvi)). Following the
  style of the function for performing screening via `limma`, users may write
  their own screening functions and are invited to contribute such functions to
  the core software package by opening pull requests at the GitHub repository.

2. Nonparametric estimates of VIMs, for the specified parameter of interest, are
  computed at each of the CpG sites passing the screening step. The VIMs are
  defined in such a way that the estimated effects is of an exposure/treatment
  on the methylation status of a target CpG site, controlling for the observed
  methylation status of the neighbors of that site. Currently, routines are
  adapted from the [`tmle`](https://CRAN.R-project.org/package=tmle) R package.
  Future releases will support doubly-robust estimates of these VIMs (via the
  [`drtmle`](https://cran.r-project.org/web/packages/drtmle/index.html) package)
  and add parameters for continuous treatments/exposures (via the
  [`tmle.npvi`](https://CRAN.R-project.org/package=tmle.npvi) package).

@@

* The _average treatment effect_ (ATE): The effect of a binary exposure or
  treatment on the observed methylation at a target CpG site is estimated,
  controlling for the observed methylation at all other CpG sites in the same
  neighborhood as the target site, based on an additive form. Often denoted
  $\psi_0 = \psi_0(1) - \psi_0(0)$, the parameter estimate represents the
  __additive difference__ in methylation that would have been observed at the
  target site had all observations received the treatment versus the
  counterfactual under which none received the treatment.

* The _relative risk_ (RR): The effect of a binary exposure or treatment on the
  observed methylation at a target CpG site is estimated, controlling for the
  observed methylation at all other CpG sites in the same neighborhood as the
  target site, based on a geometric form. Often denoted,
  $\psi_0 = \frac{\psi_0(1)}{\psi_0(0)}$, the parameter estimate represents the
  __multiplicative difference__ in methylation that would have
  been observed at the target site had all observations received the treatment
  versus the counterfactual under which none received the treatment.

Support for continuous-valued treatments or exposures is _planned but not yet
available_, though work is underway to incorporate into our methodology the
following

* A _nonparametric variable importance measure_ (NPVI) [@chambaz2012estimation]:
  The effect of continuous-valued exposure or treatment (the observed
  methylation at a target CpG site) on an outcome of interest is estimated,
  controlling for the observed methylation at all other CpG sites in the same neighborhood as the target (treatment) site, based on a parameter that
  compares values of the treatment against a reference value taken to be the
  null. In particular, the implementation provided is designed to assess the
  effect of differential methylation at the target CpG site on a (typically) phenotype-level outcome of interest (e.g., survival), in effect providing an nonparametric evaluation of the impact of methylation at the target site on
  said outcome.

We have adopted a class `methytmle` to help organize the functionality within
this package. The `methytmle` class builds upon the `GenomicRatioSet` class
provided by the `minfi` package so all of the slots of `GenomicRatioSet` are
contained in a `methytmle` object. The new class introduced in the `methyvim`
package includes several new slots:

* `call` - the form of the original call to the `methyvim` function.
* `screen_ind` - indices identifying CpG sites that pass the screening process.
* `clusters` - non-unique IDs corresponding to the manner in wich sites are
  treated as neighbors. These are assigned by genomic distance (bp) and respect
  chromosome boundaries (produced via a call to `bumphunter::clusterMaker`).
* `var_int` - the treatment/exposure status for each subject. Currently, these
  must be binary, due to the definition of the supported targeted parameters.
* `param` - the name of the target parameter from which the estimated VIMs are
  defined.
* `vim` - a table of statistical results obtained from estimating VIMs for each
  of the CpG sites that pass the screening procedure.
* `ic` - the measured array values for each of the CpG sites passing the
  screening, transformed into influence curve space based on the chosen target
  parameter.

The interested analyst might consider consulting the documentation of the
`minfi` package for an in-depth description of all of the other slots that
appear in this class [@aryee2014minfi].

We refer the reader to the package vignette, "`methyvim`: Variable Importance
Measures for Differential Methylation" hosted with the package on Bioconductor,
for a detailed description of the methods used in this package.
(NEED CITATION FOR VIGNETTE)

@@

A standard computer with the latest version of R and Bioconductor 3.6 installed
will handle applications of the `methyvim` package.

@@

### Preliminaries: Setting up the Data

@@

### Differential Methylation Analysis

@@

### Visualization of Results

@@

The purpose of such a plot is to ensure that very low (possibly statistically
significant) p-values do not arise from cases of low variance. This appears to
be the case in the plot above (notice that most parameter estimates are _near
zero_, even in cases where the raw p-values are quite low).

@@

# Summary <!-- Optional - only if NO new datasets are included -->

This section is required if the paper does not include novel data or analyses.
It allows authors to briefly summarize the key points from the article.

# Software availability
This section will be generated by the Editorial Office before publication.
Authors are asked to provide some initial information to assist the Editorial
Office, as detailed below.

1. URL link to where the software can be downloaded from or used by a non-coder (AUTHOR TO PROVIDE; optional)
2. URL link to the author's version control system repository containing the source code (AUTHOR TO PROVIDE; required)
3. Link to source code as at time of publication (*F1000Research* TO GENERATE)
4. Link to archived source code as at time of publication (*F1000Research* TO GENERATE)
5. Software license (AUTHOR TO PROVIDE; required)

@@

NH designed and implemented the tool, applied the tool to the use case, and
drafted the vignette. RP helped design the tool and drafted the article.

# Competing interests
All financial, personal, or professional competing interests for any of the
authors that could be construed to unduly influence the content of the article
must be disclosed and will be displayed alongside the article. If there are no
relevant competing interests to declare, please add the following: 'No competing
interests were disclosed'.

# Grant information
Please state who funded the work discussed in this article, whether it is your
employer, a grant funder etc. Please do not list funding that you have that is
not relevant to this specific piece of research. For each funder, please state
the funder’s name, the grant number where applicable, and the individual to whom
the grant was assigned. If your work was not funded by any grants, please
include the line: 'The author(s) declared that no grants were involved in
supporting this work.'

# Acknowledgments
This section should acknowledge anyone who contributed to the research or the
article but who does not qualify as an author based on the criteria provided
earlier (e.g. someone or an organization that provided writing assistance).
Please state how they contributed; authors should obtain permission to
acknowledge from all those mentioned in the Acknowledgments section.

Please do not list grant funding in this section.

>>>>

---
title: '`methyvim`: an R package for Variable Importance Measures for
Differential Methylation Analysis'
author:
  - name: Nima S. Hejazi
    affiliation: Address of author1
  - name: Rachael Phillips
    affiliation: orchid.org/0000-0002-8474-591X
  - name: Alan E. Hubbard
    affiliation: Address of author2
  - name: Mark J. van der Laan
    affiliation: Address of author2
abstract:
  We present a general algorithm for the nonparametric estimation of
  treatment/exposure effects on DNA methylation at CpG sites, complete with straightforward statistical inference for such estimates. This approach
  leverages variable importance measures, a class of parameters that arise in
  causal inference and are defined in such a manner that they provide targeted
  estimates of the relative importance of individual CpG site with respect to
  a binary exposure/treatment assigned at the level of subjects. Additionally,
  this procedure is computationally efficient; incorporating a preliminary
  screening step to isolate a subset of sites for which there is cursory
  evidence of differential methylation and a multiple testing correction step
  to control the False Discovery Rate as if all sites were tested. This
  technique for analysis of differentially methylated positions provides an
  avenue to move beyond linear models without the loss of interpretable
  statistical inference.

@@

DNA methylation is a fundamental epigenetic process known to play an important
role in the regulation of gene expression. DNA CpG methylation involves the
addition of a methyl group ($\text{CH}_3$) to the fifth carbon of the cytosine
ring structure to form 5-methylcytosine. Numerous biological and medical studies
have implicated DNA CpG methylation as playing a role in disease and
development [@robertson2005dna]. Perhaps unsurprisingly then, biotechnologies
have been developed to study the molecular mechanisms of this epigenetic
process. Modern assays, like the Illumina Infinium Methylation assay,
allow for quantitative interrogation of DNA methylation of CpG sites scattered
across the genome at single-nucleotide resolution; moreover, much effort
has been invested, by the bioinformatics community, in the development of tools
for properly removing technological effects that may contaminate biological
signatures measured by such assays. Despite these advances in both biological
and bioninformatical techniques, most statistical methods available for
differential analysis of data produced by such assays rely on over-simplified
(often generalized linear) models. This is because typical statistical
estimation approaches are not achievable for such high-dimensional data without restrictive modeling assumptions. When these assumptions are violated, estimates
of the population-level effect of an exposure or treatment may have large bias.
Data-adaptive estimation procedures that utilize machine learning can control
for high-dimensional confounders when estimating treatment/exposure effects.
However, interpretable statistical inference (i.e. confidence intervals,
hypothesis tests) resulting from these data-adaptive estimates is challenging [@libbrecht2015machine]. In this paper, we present an alternative such
statistical analysis approaches in the form of nonparametric estimation
procedures that provide simple and well interpretable statistical inference.
Inspired by concepts of counterfactual effects from the causal inference
literature and machine learning algorithms, we provide a computationally
efficient technique for obtaining targeted estimates of nonparametric variable
importance measures (VIM) for a pre-screened set of CpGs, controlling for
the False Discovery Rate (FDR) as if all sites were tested. Under counterfactual assumptions, these estimators have asymptotic normal sampling distributions
and thus provide reliable inference. Counterfactuals are "counter"-fact, defined
in terms of an unrealistic yet ideal experiment. In the context of DNA
methylation and for a binary exposure/treatment, we define the counterfactual
outcomes as the Beta-value/M-value a CpG site would have if all subjects were
exposed/treated and the Beta-value/M-value a CpG site would have if all
subjects were not exposed/treated. Of course, these counterfactual outcomes are
impossible to observe but they are very helpful for us to asses the importance
of a CpG site relative to a subjects treatment/exposure and they can be
statistically estimated [@van2006statistical]. Our VIM definition uses the
concepts of counterfactuals and the framework for estimation is built on
targeted maximum likelihood estimation or targeted minimum loss-based estimation
(TMLE) [@van2006targeted]. This methodology targets the individual importance
of each CpG site by taking advantage of the prediction power of machine learning
algorithms. The mathematical properties of this approach have been explored in [@tuglus2008targeted] and we focus on our contribution to it's application,
methyvim, an R package that accommodates this technique for differential
methylation analysis.

@@

The core functionality of this package is made available via the eponymous
methyvim function, which implements a statistical algorithm designed to
compute targeted estimates of VIMs, defined in such a way that the VIMs
represent parameters of scientific interest in computational biology
experiments; moreover, these VIMs are defined such that they may be estimated in
a manner that is very nearly assumption-free, that is, within a fully
nonparametric statistical model. The statistical algorithm consists in
several major steps:

1. Pre-screening of genomic sites is used to isolate a subset of sites for
  which there is cursory evidence of differential methylation. For the sake of
  computational feasibility, targeted minimum loss-based estimates of VIMs are
  computed only for this subset of sites. Currently, the available screening
  approach adapts core routines from the
  [limma](http://bioconductor.org/packages/release/bioc/html/limma.html) R
  package. Future releases will support functionality from other
  packages (e.g., [randomForest](https://cran.r-project.org/web/packages/randomForest/index.html),
  [tmle.npvi](https://CRAN.R-project.org/package=tmle.npvi)). Following the
  style of the function for performing screening via limma, users may write
  their own screening functions and are invited to contribute such functions to
  the core software package by opening pull requests at the GitHub repository.

2. Nonparametric estimates of VIMs, for the specified parameter of interest, are
  computed at each of the CpG sites passing the screening step. The VIMs are
  defined in such a way that the estimated effects is of an exposure/treatment
  on the methylation status of a target CpG site, controlling for the observed
  methylation status of the neighbors of that site. Currently, routines are
  adapted from the [tmle](https://CRAN.R-project.org/package=tmle) R package.
  Future releases will support doubly-robust estimates of these VIMs (via the
  [drtmle](https://cran.r-project.org/web/packages/drtmle/index.html) package)
  and add parameters for continuous treatments/exposures (via the
  [tmle.npvi](https://CRAN.R-project.org/package=tmle.npvi) package).

@@

* The average treatment effect (ATE): The effect of a binary exposure or
  treatment on the observed methylation at a target CpG site is estimated,
  controlling for the observed methylation at all other CpG sites in the same
  neighborhood as the target site, based on an additive form. Often denoted
  $\psi_0 = \psi_0(1) - \psi_0(0)$, the parameter estimate represents the
  additive difference in methylation that would have been observed at the
  target site had all observations received the treatment versus the
  counterfactual under which none received the treatment.

* The relative risk (RR): The effect of a binary exposure or treatment on the
  observed methylation at a target CpG site is estimated, controlling for the
  observed methylation at all other CpG sites in the same neighborhood as the
  target site, based on a geometric form. Often denoted,
  $\psi_0 = \frac{\psi_0(1)}{\psi_0(0)}$, the parameter estimate represents the
  multiplicative difference in methylation that would have
  been observed at the target site had all observations received the treatment
  versus the counterfactual under which none received the treatment.

Estimating the VIM corresponding to the parameters above, for discrete-valued
treatments or exposures, requires two separate regression steps: one for the
treatment mechanism (propensity score) and one for the outcome regression.
Technical details on the nature of these regressions are discussed in
@hernan2018causal, and details for estimating these regressions in the
framework of targeted minimum loss-based estimation are discussed in
@vdl2011targeted.

Support for continuous-valued treatments or exposures is planned but not yet
available, though work is underway to incorporate into our methodology the
following

* A nonparametric variable importance measure (NPVI) [@chambaz2012estimation]:
  The effect of continuous-valued exposure or treatment (the observed
  methylation at a target CpG site) on an outcome of interest is estimated,
  controlling for the observed methylation at all other CpG sites in the same neighborhood as the target (treatment) site, based on a parameter that
  compares values of the treatment against a reference value taken to be the
  null. In particular, the implementation provided is designed to assess the
  effect of differential methylation at the target CpG site on a (typically) phenotype-level outcome of interest (e.g., survival), in effect providing an nonparametric evaluation of the impact of methylation at the target site on
  said outcome.

#### Class methytmle

We have adopted a class methytmle to help organize the functionality within
this package. The methytmle class builds upon the GenomicRatioSet class
provided by the minfi package so all of the slots of GenomicRatioSet are
contained in a methytmle object. The new class introduced in the methyvim
package includes several new slots:

* __call__ - the form of the original call to the methyvim function.
* __screen_ind__ - indices identifying CpG sites that pass the screening process.
* __clusters__ - non-unique IDs corresponding to the manner in wich sites are
  treated as neighbors. These are assigned by genomic distance (bp) and respect
  chromosome boundaries (produced via a call to bumphunter::clusterMaker).
* __var_int__ - the treatment/exposure status for each subject. Currently, these
  must be binary, due to the definition of the supported targeted parameters.
* __param__ - the name of the target parameter from which the estimated VIMs are
  defined.
* __vim__ - a table of statistical results obtained from estimating VIMs for
  each of the CpG sites that pass the screening procedure.
* __ic__ - the measured array values for each of the CpG sites passing the
  screening, transformed into influence curve space based on the chosen target
  parameter.

We refer the reader to the package vignette, "methyvim: Variable Importance
Measures for Differential Methylation" hosted with the package on Bioconductor,
for a detailed description of the methods used in this package.
(NEED CITATION FOR VIGNETTE)

@@

A standard computer with the latest version of R and Bioconductor 3.6 installed
will handle applications of the methyvim package.

@@

#### Preliminaries: Setting up the Data

@@

#### Differential Methylation Analysis

@@

#### Visualization of Results

@@

The purpose of such a plot is to ensure that very low (possibly statistically
significant) p-values do not arise from cases of low variance. This appears to
be the case in the plot above (notice that most parameter estimates are near
zero, even in cases where the raw p-values are quite low).

@@

# Summary

Here we introduce the R package methyvim, an algorithm for differential
methylation analysis that moves beyond linear models. Additionally, this
estimation procedure produces straightforward statistical inference. Inspired by
concepts of counterfactual effects from the causal inference literature and
machine learning algorithms, we provide a computationally efficient technique
for obtaining targeted estimates of nonparametric variable importance measures
for a pre-screened set of CpGs, controlling for the False Discovery Rate as if
all sites were tested, and the estimation framework is built on targeted minimum loss-based estimation. By taking advantage of the prediction power of machine
learning algorithms, this methodology targets the individual importance of each
CpG site, providing a fully nonparametric statistical model for the analysis of
differential methylation.

# Software availability

@@

NH designed and implemented the tool, applied the tool to the use case, and
drafted the vignette. RP helped design the tool and drafted the article.
AH and ML serve as the advisors for the development of this tool.

# Competing interests

No competing interests were disclosed.

# Grant information

The author(s) declared that no grants were involved in supporting this work.
